Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-03-26 analysis/paper/paper.Rmd
Slight changes to the data prep for PCA

```{r libraries, eval = TRUE}
library(tidyverse)
library(FactoMineR)
library(factoextra)
library(patchwork)
```

@@

Figure \@ref(fig:cor) displays a correspondence analysis using the lithic count data. \@ref(fig:cor)

@@

lithics2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project) %>% 
  rowwise() %>% 
  # Create a new column with mutate() called "artefact_count" which is to hold 
  # the sum() of the artefact columns for each row. 
  mutate(
    artefact_count = sum(c_across(core:flake_retouch_other))) %>% 
  mutate(
    mean_date = mean(c(reported_start_bp, reported_end_bp)),
    .keep = "unused") %>% 
  ungroup() %>% 
  mutate(chip_freq = artefact_count / sum(chip, chip_other),
         core_freq = artefact_count / sum(core, core_other),
         primary_freq = artefact_count / sum(core, core_fragment, blade, 
                     microblade, chip, fragment, flake, flake_other, 
                     fragment_other, chip_other, blade_other,  microblade_other, 
                     core_other, core_fragment_other),
         secondary_freq = artefact_count / sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_point_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other),
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-primary_lithics, -secondary_lithics, -excavated_m3, -artefact_count) %>% 
  column_to_rownames("site")
```

```{r pca-analysis, fig.cap = "PCA.", fig.height = 7, fig.width = 10}
pcalithics <- PCA(select(lithics2, -mean_date), graph = FALSE)
fviz_pca(pcalithics, col.ind = lithics2$mean_date, col.var = "black", 
         repel = TRUE) +  scale_color_gradient2(name = "Years cal. BP", 
         low="blue", mid="yellow", high="red", midpoint = centdate)

@@

Figure \@ref(fig:pca-analysis) displays a principle components analysis using the continuous measures as defined by @Bicho2020

>>>>

```{r libraries, eval = TRUE}
library(tidyverse)
library(FactoMineR)
library(factoextra)
```

@@

Figure \@ref(fig:cor) displays a correspondence analysis using the lithic count data.

@@

lithics2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project) %>% 
  rowwise() %>% 
  # Create a new column with mutate() called "artefact_count" which is to hold 
  # the sum() of the artefact columns for each row. 
  mutate(
    artefact_count = sum(c_across(core:flake_retouch_other))) %>% 
  mutate(
    mean_date = mean(c(reported_start_bp, reported_end_bp)),
    .keep = "unused") %>% 
  ungroup() %>% 
  # The reason for not using .keep below is for the option to retain artefact_count
  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
         core_freq = sum(core, core_other) / artefact_count,
         primary_freq = sum(core, core_fragment, blade, 
                     microblade, chip, fragment, flake, flake_other, 
                     fragment_other, chip_other, blade_other,  microblade_other, 
                     core_other, core_fragment_other) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_point_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other) / artefact_count,
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-primary_lithics, -secondary_lithics, -excavated_m3, -artefact_count) %>% 
  column_to_rownames("site")
```

```{r pca-analysis, fig.cap = "PCA.", fig.height = 7, fig.width = 10}
pcalithics <- PCA(select(lithics2, -mean_date), scale.unit = TRUE, graph = FALSE)
fviz_pca(pcalithics, col.ind = lithics2$mean_date, col.var = "black", 
         repel = TRUE) +  scale_color_gradient2(name = "Years cal. BP", 
         low="blue", mid="yellow", high="red", midpoint = centdate)

@@

Figure \@ref(fig:pca-analysis) displays a principle components analysis using continuous measures as defined by @Bicho2020
